---
title: "Survival Analysis with Aster and R - Churn Use Case"
author: "Gregory Kanevsky"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document: default
  html_notebook: default
  pdf_document: default
---

```{r Setup and Functions, cache=FALSE, include=FALSE, echo=FALSE}
library(TeradataAsterR)
library(ggplot2)
library(ggthemes)
library(RColorBrewer)

createHazardRatioPlot <- function(data, cat='category', value='exp_coef', xlab=NULL,
                                  ylab = ifelse(value=='exp_coef', expression(exp(beta)), expression(beta)),
                                  title=NULL, subtitle=NULL, base_size=14, angle=270) {
  ggplot(data) +
    geom_bar(aes_string(x=cat, y=value, fill=cat), stat = "identity") +
    scale_fill_manual(values=colorRampPalette(solarized_pal()(8))(dim(data)[[1]])) +
    geom_hline(aes(yintercept=ifelse(value=='exp_coef', 1, 0)), size=0.7, linetype="42") +
    labs(title=title, subtitle=subtitle, y=ylab, x=xlab) +
    theme_minimal(base_size = base_size, base_family = 'mono') +
    theme(legend.position = "none",
          axis.text.x = element_text(angle=angle, vjust = 0))
}

conn = ta.connect("PreSalesCluster1-dallas", database = "mike")

masterView = 'p17.cust_metrics_summary_june2016'
```

## Feature Engineering for Survival Analysis
With feature engineering we accomplish several tasks:

  * Create special features for duration and event that correspond to survival length and outcome.
  * Create more robust covariates for survival analysis such as sms message count in 100s, and
    voice minutes to hours
    
```{r Feature Engineering, cache=TRUE, include=TRUE, echo=TRUE}
customer.data.ta = ta.data.frame("cust_metrics_summary_june2016", schemaName = "p17")
survival.columns = c('cust_id','plan','region','data_usage_avg','sms_usage_avg','voice_minutes_avg','cust_days',
                     'close_dt','open_dt')
survival.data.ta = ta.transform(customer.data.ta[, survival.columns],
                                voice_24hours_avg = voice_minutes_avg / (24 * 60),
                                sms100_usage_avg = sms_usage_avg / 100,
                                duration = if(!is.null(close_dt)) close_dt - open_dt
                                            else '1/31/2017' - open_dt,
                                event = if(!is.null(close_dt)) 1 else 0)

ta.dropTable("customers_for_coxph", schemaName = "p17")
ta.create(survival.data.ta[, union(setdiff(survival.columns, c('close_dt','open_dt')), 
                                   c('voice_24hours_avg','sms100_usage_avg','duration','event'))],
          table = 'customers_for_coxph', schemaName = 'p17', 
          tableType = 'fact', partitionKey = 'cust_id', row.names = TRUE,
          colTypes = c(event='int'))


customer.data.ta = ta.data.frame("customers_for_coxph", schemaName = "p17")
```

## Creating Survival Model

```{r Survival Model, cache=FALSE, include=TRUE, echo=TRUE}
all.features = setdiff(ta.colnames(customer.data.ta), 
                       c("row_names", "cust_id", "event", "cust_days", "duration",
                         "voice_minutes_avg", "sms_usage_avg"))
categorical.features = intersect(all.features, c("plan","region"))

cox.model = aa.coxph(data = customer.data.ta, 
                     feature.columns = all.features,
                     categorical.columns = categorical.features,
                     time.interval.column = "cust_days",
                     event.column = "event",
                     accumulate = "cust_id")

names(cox.model)
ta.head(cox.model$coefficient.table)
ta.head(cox.model$linear.predictor.table)

coeffs = as.data.frame(cox.model$coefficient.table)
```

## Hazard Ratios

### Hazad Ratio: Plan (Categorical)

```{r Hazard Ratio Plan, cache=FALSE, echo=TRUE, fig.width=7, fig.height=6}
data = coeffs[coeffs$predictor=='plan' &
                (coeffs$p_value <= 0.05 | is.nan(coeffs$p_value)), ]
data$category = factor(data$category, levels = data$category[order(data$exp_coef)], ordered = TRUE)
createHazardRatioPlot(data, 
                      title = expression(paste("Log Hazard Ratios ", exp(beta), " for Plans")),
                      subtitle = "Plan 'Offer 20' as baseline at 1. Only plans with significant values included")
createHazardRatioPlot(data, value = "coefficient", 
                      title = "Hazard Ratio Coefficients for Plans",
                      subtitle = "Plan 'Offer 20' as baseline at 0. Only plans with signficant values included")
```

### Hazard Ratio: Region (Categorical)

```{r Hazard Ratio Region, cache=FALSE, echo=TRUE, fig.width=7, fig.height=6}
data = coeffs[coeffs$predictor=='region' &
                (coeffs$p_value <= 0.05 | is.nan(coeffs$p_value)), ]
data$category = factor(data$category, levels = data$category[order(data$exp_coef)], ordered = TRUE)
createHazardRatioPlot(data, 
                      title = expression(paste("Log Hazard Ratio ", exp(beta), " for Regions")),
                      subtitle = "Region R1 as baseline at 1. Only regions with significant values included")
createHazardRatioPlot(data, value = 'coefficient', angle = 0,
                      title = "Hazard Ratio Coefficients for Regions",
                      subtitle = 'Region R1 as baseline at 0. Only plans with significant values included')

```

### Hazard Ratio: Usage (Numerical Features)
```{r Hazard Ratio Numerical, cache=FALSE, echo=TRUE, fig.width=7, fig.height=6}
data = coeffs[!coeffs$predictor %in% c('region','plan'), ]
data$predictor = factor(data$predictor, levels = data$predictor[order(data$exp_coef)], ordered = TRUE)
createHazardRatioPlot(data, 'predictor', 
                      title = expression(paste("Log Hazard Ratio ", exp(beta), " of Numeric Predictors")),
                      subtitle = "Voice, Data, and Text Usage Changes by One Unit", angle=0)
createHazardRatioPlot(data, 'predictor', value = 'coefficient', 
                      title = expression(paste("Hazard Ratio ", beta, " of Numeric Predictors")),
                      subtitle = "Voice, Data, and Text Usage Changes by One Unit", angle=0)
```

## Comparing Customer Survivability
```{r Customer Survival Curves, cache=FALSE, echo=TRUE, fig.width=7, fig.height=6}
survival.probs = aa.cox.survfit(object = cox.model$coefficient.table,
               cox.model.table = cox.model$linear.predictor.table,
               predict.table = ta.subset(customer.data.ta, cust_id %in% c('1050527710','1064467365')),
               predict.feature.columns = all.features,
               predict.feature.names = all.features,
               accumulate = 'cust_id')

data = as.data.frame(survival.probs$survival.probability)
data$cust_id = factor(data$cust_id)
ggplot(data) +
  geom_line(aes(time_interval, survival_prob, group=cust_id, color=cust_id), size=1) +
  scale_color_solarized(name="Customer") +
  labs(x="Days", y="Proportion Not Canceling", 
       title = "Survival Curves by Customers", subtitle = NULL) +
  theme_minimal(base_size = 14, base_family = 'mono') +
  theme(legend.position = "bottom")

```

## Hazard Ratios by Customers

### Reference Customer for Proportional Hazard
Let's evaluate hazard ratios between all customers and reference mean "customer".
This is scoring of customers using computed Cox PH model.

First, create "average" customer to use as a reference (a baseline) to compute proportional hazards. 
Such customer could be specificly designed based on research use case, e.g. analyzing certain
demographics, customer experience, etc.

```{r Typical Customer, cache=FALSE, echo=TRUE}
customer_numeric_means = ta.colMeans(customer.data.ta[,setdiff(all.features, 
                                                            categorical.features)])
customer_all_means = cbind(as.data.frame(t(customer_numeric_means)),
                           data.frame(plan="Plan 30", region="R2", 
                                      stringsAsFactors = FALSE))
ta.dropTable("customer_means", schemaName = "p17")
ta.create(data=customer_all_means, table="customer_means", schemaName = "p17", 
          tableType = "dimension")
customer.means.ta = ta.data.frame("customer_means", schemaName = "p17")
```

### Scoring Customers

```{r Scoring Customers, cache=FALSE, echo=TRUE}
# compute ratios
customer.ratios = aa.cox.predict(
  object = cox.model, 
  predicts = customer.data.ta,
  refs = customer.means.ta,
  predicts.partition.column = "1",
  predict.feature.columns = all.features,
  predict.feature.names = all.features,
  ref.feature.columns = all.features,
  refs.partition.column = c('1'),
  accumulate = c("cust_id")
  )

predictions.coxph = customer.ratios[[1]]
n = ta.dim(predictions.coxph)[[1]]
```

### Analyzing Proportional Hazard Customers

```{r CustomerPercentiles, cache=FALSE, echo=TRUE}
# Get just bottom and top N customers
N = 100
percentiles = aa.percentile(predictions.coxph, data.partition.column = "1",
                            target.columns = 'hazardratio',
                            percentile = c(N/n * 100, (1.0 - N/n)*100))
percentiles.values = as.matrix(percentiles$result)
bottom.value = percentiles.values[[1,2]]
top.value = percentiles.values[[2,2]]

bottom_and_top_customers = as.data.frame(
  ta.transform(ta.subset(predictions.coxph, hazardratio <= bottom.value | hazardratio > top.value), 
               group = if(hazardratio <= bottom.value) 'bottom' else 'top'),
  stringsAsFactors = FALSE)
bottom_and_top_customers$cust_id =
  factor(bottom_and_top_customers$cust_id, 
         levels = bottom_and_top_customers$cust_id[order(bottom_and_top_customers$hazardratio)], 
         ordered = TRUE)

bottom_and_top_customers$group = factor(bottom_and_top_customers$group, 
                                        levels=c('top','bottom'), 
                                        labels=c('High Risk','Low Risk'),
                                        ordered = TRUE)
```

### Customers with Regions

Besides each bar stands concrete customer sorted by their proportional hazard ratio.
We always select bottom (low riks) and top (high risk) to display.
```{r HazardRatiosWithRegions, echo=FALSE, fig.width=8, fig.height=6}
ggplot(bottom_and_top_customers) +
  geom_bar(aes(cust_id, hazardratio, fill=region), stat='identity') +
  facet_wrap(~group, scales = "free", ncol=1) +
  scale_fill_solarized(guide=guide_legend(title="Regions", nrow=1)) +
  labs(x="Customer IDs", y="Hazard Ratio", 
       title=paste("Bottom and Top",N,"Customers by Risk"), 
       subtitle="by proportional hazard ratio to 'average' customer") +
  theme_tufte(ticks=FALSE, base_size = 14, base_family = 'mono') + 
  theme(legend.position = "bottom",
        axis.text.x = element_blank())
```

### Customers with Plans

```{r HazardRatiosWithPlans, echo=FALSE, fig.width=8, fig.height=6}
ggplot(bottom_and_top_customers) +
  geom_bar(aes(cust_id, hazardratio, fill=plan), stat='identity') +
  facet_wrap(~group, scales = "free", ncol=1) +
  scale_fill_solarized(guide=guide_legend(title="Plans", nrow=2)) +
  labs(x="Customer IDs", y="Hazard Ratio", 
       title=paste("Bottom and Top",N,"Customers by Risk"), 
       subtitle="by proportional hazard ratio to 'average' customer") +
  theme_tufte(ticks=FALSE, base_size = 14, base_family = 'mono') + 
  theme(legend.position = "bottom",
        axis.text.x = element_blank())
```

### Customers with Phone Usages

```{r HazardRatiosWithData, echo=FALSE, fig.width=8, fig.height=6}
ggplot(bottom_and_top_customers) +
  geom_bar(aes(cust_id, hazardratio, fill=data_usage_avg), stat='identity') +
  facet_wrap(~group, scales = "free", ncol=1) +
  scale_fill_gradient(low = "#56B1F7", high = "#132B43", guide = guide_colorbar(title='Data Usage (GB)')) +
  labs(x="Customer IDs", y="Hazard Ratio", 
       title=paste("Bottom and Top",N,"Customers by Risk"), 
       subtitle="by proportional hazard ratio to 'average' customer") +
  theme_tufte(ticks=FALSE, base_size = 14, base_family = 'mono') + 
  theme(legend.position = "bottom",
        axis.text.x = element_blank())
```

```{r HazardRatiosWithSMS, echo=FALSE, fig.width=8, fig.height=6}
ggplot(bottom_and_top_customers) +
  geom_bar(aes(cust_id, hazardratio, fill=sms100_usage_avg), stat='identity') +
  facet_wrap(~group, scales = "free", ncol=1) +
  scale_fill_gradient(low = "#56B1F7", high = "#132B43", guide = guide_colorbar(title='SMS Usage (100s)')) +
  labs(x="Customer IDs", y="Hazard Ratio", 
       title=paste("Bottom and Top",N,"Customers by Risk"), 
       subtitle="by proportional hazard ratio to 'average' customer") +
  theme_tufte(ticks=FALSE, base_size = 14, base_family = 'mono') + 
  theme(legend.position = "bottom",
        axis.text.x = element_blank())
```

```{r HazardRatiosWithVoice, echo=FALSE, fig.width=8, fig.height=6}
ggplot(bottom_and_top_customers) +
  geom_bar(aes(cust_id, hazardratio, fill=voice_24hours_avg), stat='identity') +
  facet_wrap(~group, scales = "free", ncol=1) +
  scale_fill_gradient(low = "#56B1F7", high = "#132B43", guide = guide_colorbar(title='Voice Usage (24hrs)')) +
  labs(x="Customer IDs", y="Hazard Ratio", 
       title=paste("Bottom and Top",N,"Customers by Risk"), 
       subtitle="by proportional hazard ratio to 'average' customer") +
  theme_tufte(ticks=FALSE, base_size = 14, base_family = 'mono') + 
  theme(legend.position = "bottom",
        axis.text.x = element_blank())

```