---
title: "K-Means Clustering Workflow with Aster and R - Churn Use Case"
author: "Gregory Kanevsky"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document: default
  html_notebook: default
  pdf_document: default
---

```{r Setup and Functions, cache=FALSE, include=FALSE, echo=FALSE}
library(TeradataAsterR)
library(toaster)
library(ggplot2)
library(scales)
library(ggthemes)

conn = ta.connect("Aster", database = "demo3")

masterView = 'p17.cust_metrics_summary_june2016'
```

# Determining Optimal Number of Clusters

## Running K-means and Varying K (Number of Clusters)
By varying k from 2 to K=20 and saving models we can track how k-means relevant metrics change.
```{r DetermineNumberOfClusters, cache=TRUE, echo=TRUE}

K = 20
kmeansModels = list()
for(k in 2:K) {
  kmeansCustomers1 = computeKmeans(taConnection, "p17.cust_metrics_summary_june2016", centers=k, 
                        include=c('voice_minutes_avg', 'sms_usage_avg', 'data_usage_avg'),
                        scale = TRUE, id='cust_id', idAlias = "cust_id", 
                        where="plan IN ('Plan 40','Plan 30','Plan 35','Plan 39','Plan 25','Plan 50')",
                        scaledTableName=paste0('kmeans_customer_scaled_k_',k), 
                        centroidTableName=paste0('kmeans_customer_centroids_k_',k), schema='p17',
                        iterMax = 500)
  kmeansModels[[k-1]] = kmeansCustomers1
}

totalRunningTimeMinutes = sum(sapply(kmeansModels, FUN = function(x){x$time['elapsed']}))/60
```

## Plot Total Withinness Trend
```{r PlotWithinness, cache=FALSE, echo=TRUE, fig.width=7, fig.height=6}
# Graph for Total Withinness
data = data.frame(K=integer(0), totwithinss=numeric(0))
for(n in 2:(length(kmeansModels)+1)) {
  kmeansModel = kmeansModels[[n-1]]
  data = rbind(data,
               data.frame(K=n, totwithinss=kmeansModel$tot.withinss))
}
ggplot(data) +
  geom_line(aes(K, totwithinss), size=1) + geom_point(aes(K, totwithinss), size=2) + 
  labs(title="Within groups Sums of Squares vs. the Number of Clusters", 
       x="Number of Clusters", y="Within groups Sums of Squares") +
  scale_x_continuous(breaks=seq(0,K,2)) + 
  scale_y_continuous(labels = comma) +
  theme_minimal(base_size = 12, base_family = 'serif')
```

## Picking the Model
We pick 6 cluster model based on the elbow position:
```{r PickModel, cache=TRUE, echo=TRUE}

k = 6
kmeans.model = computeKmeans(taConnection, "p17.cust_metrics_summary_june2016", centers=k, 
                        include=c('voice_minutes_avg', 'sms_usage_avg', 'data_usage_avg'),
                        scale = TRUE, 
                        aggregates = c("COUNT(*) cnt", "AVG(age) avg_age",
                                       "SUM(CASE WHEN plan = 'Plan 40' THEN 1 ELSE 0 END) plan40_count",
                                       "SUM(CASE WHEN plan = 'Plan 30' THEN 1 ELSE 0 END) plan30_count",
                                       "SUM(CASE WHEN plan = 'Plan 35' THEN 1 ELSE 0 END) plan35_count",
                                       "SUM(CASE WHEN plan = 'Plan 39' THEN 1 ELSE 0 END) plan39_count",
                                       "SUM(CASE WHEN plan = 'Plan 25' THEN 1 ELSE 0 END) plan25_count",
                                       "SUM(CASE WHEN plan = 'Plan 50' THEN 1 ELSE 0 END) plan50_count"),
                        id='cust_id', idAlias = "cust_id", 
                        where="plan IN ('Plan 40','Plan 30','Plan 35','Plan 39','Plan 25','Plan 50')",
                        scaledTableName=paste0('kmeans_picked_customer_scaled_k_',k), 
                        centroidTableName=paste0('kmeans_picked_customer_centroids_k_',k), schema='p17',
                        iterMax = 500)
```

# Reviewing Model

## Visualizing Centroids
```{r VisualizingCentroids, echo=TRUE, cache=FALSE}
createCentroidPlot(kmeans.model, format="line", groupByCluster=TRUE, coordFlip = FALSE,
                   title = "Mobile Customer Clustering: Centroids")
createCentroidPlot(kmeans.model, format="line", groupByCluster=FALSE, coordFlip = FALSE,
                   title = "Mobile Customer Clustering: Centroids")
```

## Visualizing Cluster Properties
```{r VisualizingClusters}
createClusterPlot(kmeans.model)
```

## Visualizing Cluster Samples
```{r ClusterSamples, cache=TRUE, echo=TRUE, warning=FALSE, message=FALSE}

kmeans.model = computeClusterSample(taConnection, kmeans.model, 0.05)
createClusterPairsPlot(kmeans.model)
```

# Eliminating Inactive Users from the Analysis

## Feature Distributions

### Average Monthly Voice Usage

```{r TableInfoForHist, cache=FALSE, echo=FALSE}
tableInfo = getTableSummary(taConnection, tableName = "p17.cust_metrics_summary_june2016", 
                            include=c('voice_minutes_avg', 'sms_usage_avg', 'data_usage_avg'))
```

```{r FeatureDistsVoice, cache=TRUE, echo=TRUE, fig.width=9}

voiceMinutesHist = computeHistogram(taConnection, tableName = "p17.cust_metrics_summary_june2016",
                                    columnName = "voice_minutes_avg", tableInfo = tableInfo,
                                    numbins = 80, startvalue = 0, endvalue = 80)
createHistogram(voiceMinutesHist,
                title="Monthly Average Voice Usage", xlab="Minutes", ylab="Count",
                baseSize = 12, baseFamily = 'serif')

```

### Average Monthly SMS Usage

```{r FeatureDistsSms, cache=TRUE, echo=TRUE, fig.width=9}

smsHist = computeHistogram(taConnection, tableName = "p17.cust_metrics_summary_june2016",
                                    columnName = "sms_usage_avg", tableInfo = tableInfo,
                                    numbins = 80, startvalue = 0, endvalue = 80)
createHistogram(smsHist,
                title="Monthly Average SMS Usage", xlab="Tex Message Count", ylab="Count",
                baseSize = 12, baseFamily = 'serif')

```

```{r FeatureDistsData, cache=TRUE, echo=TRUE, fig.width=9}

dataHist = computeHistogram(taConnection, tableName = "p17.cust_metrics_summary_june2016",
                                    columnName = "data_usage_avg", tableInfo = tableInfo,
                                    numbins = 50, startvalue = 0, endvalue = 10)
createHistogram(dataHist,
                title="Monthly Average Data Usage", xlab="Minutes", ylab="Count",
                baseSize = 12, baseFamily = 'serif')

```

## New Model Without Inactive Users

### Clustering Active Customers Only

We reduce number of clusters by 1 since it new data set will not contain inactive users that occupied most populated cluster before.
Note augmented *where* clause argument that filters out customers with low usage for all 3 parameters.
```{r ActiveOnlyCustomersKmeans, cache=TRUE, echo=TRUE}

k = 5
kmeansActiveModel = computeKmeans(taConnection, "p17.cust_metrics_summary_june2016", centers=k, 
                        include=c('voice_minutes_avg', 'sms_usage_avg', 'data_usage_avg'),
                        scale = TRUE, 
                        aggregates = c("COUNT(*) cnt", "AVG(age) avg_age",
                                       "SUM(CASE WHEN plan = 'Plan 40' THEN 1 ELSE 0 END) plan40_count",
                                       "SUM(CASE WHEN plan = 'Plan 30' THEN 1 ELSE 0 END) plan30_count",
                                       "SUM(CASE WHEN plan = 'Plan 35' THEN 1 ELSE 0 END) plan35_count",
                                       "SUM(CASE WHEN plan = 'Plan 39' THEN 1 ELSE 0 END) plan39_count",
                                       "SUM(CASE WHEN plan = 'Plan 25' THEN 1 ELSE 0 END) plan25_count",
                                       "SUM(CASE WHEN plan = 'Plan 50' THEN 1 ELSE 0 END) plan50_count"),
                        id='cust_id', idAlias = "cust_id", 
                        where="plan IN ('Plan 40','Plan 30','Plan 35','Plan 39','Plan 25','Plan 50') 
                              AND voice_minutes_avg >= 10 AND sms_usage_avg >= 10 AND data_usage_avg >= 2",
                        scaledTableName=paste0('kmeans_customer_active_scaled_k_',k), 
                        centroidTableName=paste0('kmeans_customer_active_centroids_k_',k), schema='p17',
                        iterMax = 500)

```

### Visualizing Centroids
```{r VisualizingCentroidsActiveCust, echo=TRUE, cache=FALSE}
createCentroidPlot(kmeansActiveModel, format="line", groupByCluster=TRUE, coordFlip = FALSE,
                   title = "Mobile Customer Clustering: Centroids")
createCentroidPlot(kmeansActiveModel, format="line", groupByCluster=FALSE, coordFlip = FALSE,
                   title = "Mobile Customer Clustering: Centroids")
```

### Visualizing Cluster Properties
```{r VisualizingClustersActiveCust}
createClusterPlot(kmeansActiveModel)
```

### Visualizing Cluster Samples
```{r ClusterSamplesActiveCust, cache=TRUE, echo=TRUE, warning=FALSE, message=FALSE}

kmeansActiveModel = computeClusterSample(taConnection, kmeansActiveModel, 0.05)
createClusterPairsPlot(kmeansActiveModel)
```
